### 2021-06-24 thesis/tex/introduction.tex
update

\chapter{Introduction}
There is no reason for us to believe that our current knowledge in particle physics is complete.
The most successful framework -- the Standard Model -- has notable shortcomings. Neither does it 
explain why there is something rather than nothing, nor does it give a candidate for dark matter\footnote{In the Standard Model,
matter and antimatter come in pairs of one particle and one antiparticle. When interacting,
antimatter \emph{annihilates} the matter, resulting in pure light. Since the universe exists,
there must have been more matter than antimatter at the Big Bang. %TODO: check and make better
Dark matter is heavy matter which does not interact with electromagnetic radiation.
No such particle exists in the Standard Model.}.

To make matters even worse, three of the massive particles are massless in the Standard Model.
In this thesis, we focus on the said particles -- the neutrinos -- and propose extensions to the Standard Model related to them. 
We then compare the effects of these extensions with collected data to see if our new theory more closely describes Nature or not.

>>>>

\chapter{Introduction}
There is no reason for us to believe that our current knowledge in particle physics is complete.
The most successful framework -- the Standard Model -- has notable shortcomings. Neither does it 
explain why there is something rather than nothing, nor does it give a candidate for dark matter\footnote{In the Standard Model,
matter and antimatter come in pairs of one particle and one antiparticle. When interacting,
antimatter \emph{annihilates} the matter, resulting in pure light. Since the universe exists,
there must have been more matter than antimatter at the Big Bang. This baryon asymmetry is unexplained in the Standard 
Model.
Dark matter is heavy matter which does not interact with electromagnetic radiation.
No such particle exists in the Standard Model.}.
To make matters even worse, three of the massless particles have been observed to have mass.
In this thesis, we focus on the said particles -- the neutrinos -- and propose extensions to the Standard Model related to them. 
We then compare the effects of these extensions with collected data to see if our new theory more closely describes Nature or not.

### 2021-06-24 thesis/tex/oscillations.tex
update

\section{Neutrino Mixing}\label{ch:oscillation}
In 1998, the Super-Kamiokande collaboration detected a $\nm$ deficit that could not be explained by any other mechanism
than the conversion muon neutrinos to electron neutrinos. This was the first conclusive evidence of a $\nm \to \ne$ oscillation.
The most interesting conclusion from the observation of neutrino flavor oscillations is the equivalence between oscillations and neutrino mass, 
with the latter being identically zero acording to the Standard Model. Thus, we are required to extend the Standard Model to incorporate neutrino 
masses and hence, neutrino oscillations.

@@

The solution to Eq.~\ref{eq:TDSE} gives ut the time evolution
in the form of plane wave solutions: 
\begin{align}
    \ket{\nu_k(t)} = e^{-iE_kt}\ket{\nu_k}\,.
\end{align}
Inserting the plane wave solution into Eq.~\ref{eq:osc_1}, we get 
\begin{align}\label{eq:nat}
    \ket{\na(t)} = \sum_k U^\dagger_{\a k} e^{-iE_kt}\ket{\nu_k}\,.
\end{align}
Now we know how to evolve and combine the mass eigenstates to form a flavor eigenstate, but how about the reverse?
 We swap the index $k\to j$ in Eq.~\ref{eq:osc_1} and multiply by $U_{\a k}$:
\begin{align}\label{eq:osc_k}
    \sum_\a U_{\a k}\ket{\na} &= \sum_{\alpha, j} U_{\a k} U_{\a j}^\dagger \ket{\nu_j} \nonumber \\
    \sum_\a U_{\a k}\ket{\na} &= \sum_{j} \delta_{kj} \ket{\nu_j} \nonumber \\
    \sum_\a U_{\a k}\ket{\na} &= \ket{\nu_k}\,,
\end{align}
where we have used the unitarity of the leptonic mixing matrix. Eqs.~\ref{eq:osc_1} and \ref{eq:nat} yield 
\begin{align}
    \ket{\na(t)} &= \sum_k U^\dagger_{\a k} e^{-iE_kt}\ket{\nu_k} \nonumber \\
    \ket{\na(t)} &= \sum_k U^\dagger_{\a k} e^{-iE_kt}\left(\sum_\b U_{\b k}\ket{\nb}\right) \nonumber \\
    \ket{\na(t)} &= \sum_{k,\b} U^\dagger_{\a k}U_{\b k} e^{-iE_kt} \ket{\nb}\,.
\end{align}
The probability of the flavor transition $\na \to \nb$ at time $t$ is $\abs{\bra{\nb}\ket{\na (t)}}^2$:
\begin{align}
    P_{\na \to \nb}(t) = \sum_{k,j} U^\dagger_{\a k}U_{\b k}U^*_{\b j}U_{\a j} e^{-i(E_k-E_j)t}\,.
\end{align}

We assume the neutrino masses $m_k$ to be extremely small compared to their associated energies $E_k$. 
Thus, $v\sim 1$, and $\abs{\vec{p}} \sim E$ making the energy-dispersion relation of Eq.~\ref{eq:disp} to first order:
\begin{align}\label{eq:ultra_rel}
    E_k &= \sqrt{\vec{p}^2 + m_k^2} \nonumber \\
        &= \vec{p}^2\sqrt{1 + \frac{m_k^2}{\vec{p}^2}} \nonumber \\
        &\approx E + \frac{m_k^2}{2E}
\end{align}
Hence, the exponential can be simplified, and simplifying the notation $P_{\na \to \nb}(t) \to P_{\ab}(t)$ we get 
\begin{align}
    P_{\ab}(t) = \sum_{k,j} U^\dagger_{\a k}U_{\b k}U^\dagger_{\b j}U_{\a j} e^{-i(m_k^2-m_j^2)t/2E}\,.
\end{align}
Now, our approximation $v\approx 1$ implies $x\approx t $, thus
\begin{align}
    P_{\ab}(x) &= \sum_{k,j} U^\dagger_{\a k}U_{\b k}U^\dagger_{\b j}U_{\a j} e^{-i(m_k^2-m_j^2)x/2E} \nonumber \\
                       &= \sum_{k,j} U^\dagger_{\a k}U_{\b k}U^\dagger_{\b j}U_{\a j} \exp(-i\frac{\dm_{kj}x}{2E})\,,
\end{align}
where we in the last step have defined the \emph{mass squared-difference} $\dm_{kj} = m_k^2 - m_j^2$. Since the oscillation probability depends on this quantity
rather than the individual masses, it is impossible
to measure the mass $m_k$ through neutrino oscillations.
Squaring the unitarity condition $\sum_k U_{\a k}U_{\b k}^\dagger = \delta_{\a \b}$ yields 
\begin{align}
    \sum_k \abs{U_{\a k}}^2\abs{U_{\b k}}^2 = \delta_{\a \b} - 2 \sum_{k>j}\Re[U_{\a k}^\dagger U_{\b k} U_{\a j} U_{\b j}^\dagger]
\end{align}
Since we take the PMNS matrix to be real and thus Hermitian, we drop the complex transpose in the following calculation. We have
\begin{align}\label{eq:Pab}%TODO: shorten?
    P_{\ab} &= \sum_{k,j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \exp(-i\frac{\dm_{kj}x}{2E}) \nonumber \\
            &=  \sum_k \abs{U_{\a k}}^2\abs{U_{\b k}}^2 + \sum_{k \neq j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \exp(-i\frac{\dm_{kj}x}{2E}) \nonumber \\
            &= \delta_{\a \b} - 2 \sum_{k>j}U_{\a k} U_{\b k} U_{\b j} U_{\a j} +\sum_{k \neq j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \exp(-i\frac{\dm_{kj}x}{2E}) \nonumber \\
            &= \delta_{\a \b} - 2 \sum_{k>j}U_{\a k} U_{\b k} U_{\b j} U_{\a j} + 2\sum_{k > j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \exp(-i\frac{\dm_{kj}x}{2E}) \nonumber \\
            &= \delta_{\a \b} - 2\sum_{k > j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \left[1-\exp(-i\frac{\dm_{kj}x}{2E})\right] \nonumber \\
            &= \delta_{\a \b} - 2\sum_{k > j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \left[1-\cos(\frac{\dm_{kj}x}{2E})\right] \nonumber \\
            &= \delta_{\a \b} - 2\sum_{k > j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \sin^2{\left(\frac{\dm_{kj}x}{4E}\right)}\,,
\end{align}
which is the probability of neutrino vacuum oscillations. %TODO: maybe plot?
The calculation for antineutrinos yield the same result if one continues to assume the realness of the mixing matrix. 
In Fig.~\ref{fig:vac_osc}, we show the energy spectra of these oscillations in the low \si{\GeV} range for neutrinos that travel \SI{12000}{\km}\footnote{This baseline
is approximately the Earth diameter, and will be the maximum travel distance which we will use later on.}.
In the left-most panel, we see a supressed $\nu_\mu \to e$ transition. We can understand this behavior by studying Eq.~\ref{eq:Pab} and the 
numerical values of the mixing matrix in Eq.~\ref{eq:Uvalues}.
\si{\GeV} oscillations are driven by the $\dm[31]$ mass-squared difference. Hence, the most important mixing matrix 
elements for $P_{\a\b}$ are $U_{\a3}, U_{\a1}$ and $U_{\b3}, U_{\b1}$. The smallness of $U_{e3}$ then supresses the
amplitude of $\Pme$.
\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{figures/vac_osc.pdf}
    \caption{$\nm$ vacuum oscillations from Eq.~\ref{eq:Pab}}\label{fig:vac_osc}
\end{figure}

@@

We start with the effective Hamiltonian for the CC process. The Feynman rules for the left panel give us 
\begin{align}
    H_{CC} = \frac{G_F}{\sqrt{2}}\left[ \bar\ne \gamma^\rho\, (1- \gamma^5)\, e \right] \left[\bar e \,\gamma_\rho\, (1- \gamma^5)\, \ne \right]
\end{align}
By using the Fierz transformation 
\begin{align}
    \mathcal{L}^{V-A}(\psi_1,\psi_2,\psi_3,\psi_4) = \mathcal{L}^{V-A}(\psi_1,\psi_4,\psi_3,\psi_2)\,,
\end{align}
we can permute the terms inside the brackets, yielding
\begin{align}\label{eq:H_fierz}
    H_{CC} = \frac{G_F}{\sqrt{2}}\left[ \bar\ne \gamma^\rho\, (1- \gamma^5)\, \ne \right] \left[\bar e \,\gamma_\rho\, (1- \gamma^5)\, e \right]\,.
\end{align}
Now, lets consider a finite volume $V$ with electron states defined as 
\begin{align}\label{eq:e_states}
    \ket{e(p_e,h_e)} = \frac{1}{2E_eV}a_e^{(h_e)\dagger}(p_e)\ket{0}\,,
\end{align}
i.e. using the creation operator $a_e^{(h_e)\dagger}(p_e)$ to create electron states from vacuum with momenta $p_e$, energy $E_e$, and helicity $h_e$.
The density distribution of electrons in $V$ is $f(E_e,T)$, which we normalize to the total number of electrons as we integrate out the momenta $p_e$:
\begin{align}\label{eq:e_density}
    \int \dd p_e^3 f(E_e,T) = N_e V = n_e
\end{align}
Here, the electron density $N_e$ will ultimately determine the strength of the effective matter potential. 
To obtain the average effective Hamiltonian, project it on the electron states in Eq.~\ref{eq:e_states} and integrate over the density and sum over the helicities:

\begin{align}\label{eq:avg_H1}
    \avg{H_{CC}} &= \int \dd p_e^3 \bra{e(p_e,h_e)} \times \frac{1}{2} \sum_{h_e} H   f(E_e,T) \ket{e(p_e,h_e)} \nonumber \\
           &= \frac{G_F}{\sqrt{2}} \int \dd p_e^3 \bra{e(p_e,h_e)} \left[ \bar\ne \gamma^\rho\, (1- \gamma^5)\, \ne \right]  f(E_e,T) \times \frac{1}{2} \sum_{h_e} \left[\bar e(x) \,\gamma_\rho\, (1- \gamma^5)\, e(x) \right] \ket{e(p_e,h_e)} \nonumber \\
           &= \frac{G_F}{\sqrt{2}} \bar\ne \gamma^\rho\, (1- \gamma^5)\, \ne \int \dd p_e^3  f(E_e,T) \times \frac{1}{2} \sum_{h_e} \bra{e(p_e,h_e)} \bar e(x) \,\gamma_\rho\, (1- \gamma^5)\, e(x)   \ket{e(p_e,h_e)}\,.
\end{align}
First, calculate the sum using trace technology
\begin{align}\label{eq:helicity_sum}
    \frac{1}{2} \sum_{h_e} \bra{e(p_e,h_e)} \bar e(x) \,\gamma_\rho\, (1- \gamma^5)\, e(x)   \ket{e(p_e,h_e)} &= \frac{1}{4E_e V} \sum_{h_e} \bar{u}_e^{h_e}(p_e) \,\gamma_\rho\, (1- \gamma^5)\, u_e^{h_e}(p_e) \nonumber \\
    &= \frac{1}{4 E_e V} \Trace{\left[ \sum_{h_e} \bar{u}_e^{h_e}(p_e) u_e^{h_e}(p_e) \,\gamma_\rho\, (1- \gamma^5)\, \right]} \nonumber \\
    &= \frac{1}{4 E_e V} \Trace{\left[ (\slashed{p}_e + m_e ) \,\gamma_\rho\, (1- \gamma^5)\, \right]} \nonumber \\
    &= \frac{(p_e)_\rho}{E_e V}\,.
\end{align}
Eq.~\ref{eq:avg_H1} now becomes 
\begin{align}\label{eq:avg_H2}
    \avg{H_{CC}} &= \frac{G_F}{\sqrt{2}E_e V} \bar\ne \, (1- \gamma^5)\, \ne \int \dd p_e^3\, \slashed{p}_e\, f(E_e,T)\,.
\end{align}
Expand the integral, and use the fact that $\vec{p}_e$ is odd:
\begin{align}
    \int \dd p_e^3\, \slashed{p}_e\, f(E_e,T) &= \int \dd p_e^3\, f(E_e,T) (\gamma^0 E_e - \vec{p}_e \cdot \vec{\gamma}) \nonumber \\
                                              &= \int \dd p_e^3\, f(E_e,T) \gamma^0 E_e \nonumber \\
                                              &= \gamma_0 E_e N_e V\,.
\end{align}
Inserting this into Eq.~\ref{eq:avg_H2}, we have
\begin{align}
    \avg{H_{CC}} &= \frac{G_F N_e}{\sqrt{2}} \bar{\nu}_e \, (1- \gamma^5)\, \ne \gamma_0 \nonumber \\
            &= \sqrt{2}G_F N_e \bar{\nu}_{eE} \gamma^0 \nu_{eL}\,,
\end{align}
where the projection operator $(1- \gamma^5)$ in the first line ensures that only the left-hand compontent of the neutrino fields interact. Thus,
\begin{align}\label{eq:V_CC1}
    V_{CC} = \sqrt{2}G_F N_e\,, \quad H_{CC}\ket{\nu_k} = V_{CC}\ket{\nu_k}\,.
\end{align}
Here we see a crucial difference between the eigenvectors between the vacuum Hamiltonian defined in Eq.~\ref{eq:osc_1} and $H_{CC}$, namely
that the CC interactions happen in the flavor basis rather than in the mass basis\footnote{A similar calculation for the NC reveals the same fact:
neutrinos interact in the flavor basis.}. In other words,
neutrinos propagate in their mass eigenstates, but interact in their flavor eigenstate. The mixing of mass eigenstates during propagation
determines if the flavor eigenstate has oscillated or not. The interactions still conserve lepton number, but the kinetic Lagrangian will not.

@@

\subsection{Matter Oscillations}
Since only $\ne$ undergo CC interactions in Earth-like matter, the $V_{CC}$ potential is zero for all other flavors. However, since all flavors undergo NC interactions the total matter potential in matrix form is
\begin{align}\label{eq:V_matrix}
    V = \begin{bmatrix}
        V_{CC} + V_{NC} & 0 & 0 \\
        0 & V_{NC} & 0 \\
        0 & 0 & V_{NC} \\
    \end{bmatrix} = V_{CC}\, \delta_{\alpha e} + V_{NC}\,.
\end{align}
Just as in Eq.~\ref{eq:TDSE}, we start with a Hamiltonian that solves the time-dependent Schrödinger equation. This time, let the Hamiltonian be 
\begin{align}
    H = H_0 + H_{I}\,,
\end{align}
where $H_0$ is the Hamiltonian in vacuum, and $H_{I}$ is our interaction Hamiltonian associated with our matter potentials.
Let the wavefunction that describes the $\nu_\a \to \nu_\b$ transition be 
\begin{align}
    \bra{\nu_\b}\ket{\nu_\a (t)}\,,
\end{align}
i.e.~the evolution of the state of a neutrino emitted at $t =0$ with flavor $\alpha$ to flavor $\beta$ at time $t$.

>>>>

\section{Neutrino Mixing}\label{ch:oscillation}
In 1998, the Super-Kamiokande collaboration detected a $\nm$ deficit that could not be explained by any other mechanism
than the conversion of muon neutrinos to electron neutrinos. This was the first conclusive evidence of a $\nm \to \ne$ oscillation.
The most interesting conclusion from the observation of neutrino flavor oscillations is the equivalence between oscillations and neutrino mass, 
with the latter being identically zero acording to the Standard Model. Thus, we are required to extend the Standard Model to incorporate neutrino 
masses before we discuss the neutrino oscillations themselves.

@@

The solution to Eq.~\ref{eq:TDSE} gives ut the time evolution
in the form of plane wave solutions: 
\begin{align}
    \ket{\nu_k(t)} = e^{-iE_kt}\ket{\nu_k}\,.
\end{align}
Inserting the plane wave solution into Eq.~\ref{eq:osc_1}, we get 
\begin{align}\label{eq:nat}
    \ket{\na(t)} = \sum_k U^\dagger_{\a k} e^{-iE_kt}\ket{\nu_k}\,.
\end{align}
Now we know how to evolve and combine the mass eigenstates to form a flavor eigenstate, but how about the reverse?
 We swap the index $k\to j$ in Eq.~\ref{eq:osc_1} and multiply by $U_{\a k}$:
\begin{align}\label{eq:osc_k}
    \sum_\a U_{\a k}\ket{\na} &= \sum_{\alpha, j} U_{\a k} U_{\a j}^\dagger \ket{\nu_j} \nonumber \\
                              &= \sum_{j} \delta_{kj} \ket{\nu_j} \nonumber \\
                              &= \ket{\nu_k}\,,
\end{align}
where we have used the unitarity of the leptonic mixing matrix. Eqs.~\ref{eq:osc_1} and \ref{eq:nat} yield 
\begin{align}
    \ket{\na(t)} &= \sum_k U^\dagger_{\a k} e^{-iE_kt}\ket{\nu_k} \nonumber \\
                 &= \sum_k U^\dagger_{\a k} e^{-iE_kt}\left(\sum_\b U_{\b k}\ket{\nb}\right) \nonumber \\
                 &= \sum_{k,\b} U^\dagger_{\a k}U_{\b k} e^{-iE_kt} \ket{\nb}\,.
\end{align}
The probability of the flavor transition $\na \to \nb$ at time $t$ is $\abs{\bra{\nb}\ket{\na (t)}}^2$:
\begin{align}
    P_{\na \to \nb}(t) = \sum_{k,j} U^\dagger_{\a k}U_{\b k}U^*_{\b j}U_{\a j} e^{-i(E_k-E_j)t}\,.
\end{align}

We assume the neutrino masses $m_k$ to be extremely small compared to their associated energies $E_k$. 
Thus, $v\sim 1$, and $\abs{\vec{p}} \sim E$ making the energy-dispersion relation of Eq.~\ref{eq:disp} to first order:
\begin{align}\label{eq:ultra_rel}
    E_k &= \sqrt{\vec{p}^2 + m_k^2} \nonumber \\
        &= \vec{p}^2\sqrt{1 + \frac{m_k^2}{\vec{p}^2}} \nonumber \\
        &\approx E + \frac{m_k^2}{2E}
\end{align}
Hence, the exponential can be simplified, and simplifying the notation $P_{\na \to \nb}(t) \to P_{\ab}(t)$ we get 
\begin{align}
    P_{\ab}(t) = \sum_{k,j} U^\dagger_{\a k}U_{\b k}U^\dagger_{\b j}U_{\a j} e^{-i(m_k^2-m_j^2)t/2E}\,.
\end{align}
Now, our approximation $v\approx 1$ implies $x\approx t $, thus
\begin{align}
    P_{\ab}(x) &= \sum_{k,j} U^\dagger_{\a k}U_{\b k}U^\dagger_{\b j}U_{\a j} e^{-i(m_k^2-m_j^2)x/2E} \nonumber \\
                       &= \sum_{k,j} U^\dagger_{\a k}U_{\b k}U^\dagger_{\b j}U_{\a j} \exp(-i\frac{\dm_{kj}x}{2E})\,,
\end{align}
where we in the last step have defined the \emph{mass squared-difference} $\dm_{kj} = m_k^2 - m_j^2$. Since the oscillation probability depends on this quantity
rather than the individual masses, it is impossible
to measure the mass $m_k$ through neutrino oscillations.
Squaring the unitarity condition $\sum_k U_{\a k}U_{\b k}^\dagger = \delta_{\a \b}$ yields 
\begin{align}
    \sum_k \abs{U_{\a k}}^2\abs{U_{\b k}}^2 = \delta_{\a \b} - 2 \sum_{k>j}\Re[U_{\a k}^\dagger U_{\b k} U_{\a j} U_{\b j}^\dagger]
\end{align}
Since we take the PMNS matrix to be real and thus Hermitian, we drop the complex transpose in the following calculation. We have
\begin{align}\label{eq:Pab}%TODO: shorten?
    P_{\ab} &= \sum_{k,j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \exp(-i\frac{\dm_{kj}x}{2E}) \nonumber \\
            &=  \sum_k \abs{U_{\a k}}^2\abs{U_{\b k}}^2 + \sum_{k \neq j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \exp(-i\frac{\dm_{kj}x}{2E}) \nonumber \\
            &= \delta_{\a \b} - 2 \sum_{k>j}U_{\a k} U_{\b k} U_{\b j} U_{\a j} +\sum_{k \neq j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \exp(-i\frac{\dm_{kj}x}{2E}) \nonumber \\
            &= \delta_{\a \b} - 2 \sum_{k>j}U_{\a k} U_{\b k} U_{\b j} U_{\a j} + 2\sum_{k > j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \exp(-i\frac{\dm_{kj}x}{2E}) \nonumber \\
            &= \delta_{\a \b} - 2\sum_{k > j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \left[1-\exp(-i\frac{\dm_{kj}x}{2E})\right] \nonumber \\
            %&= \delta_{\a \b} - 2\sum_{k > j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \left[1-\cos(\frac{\dm_{kj}x}{2E})\right] \nonumber \\
            &= \delta_{\a \b} - 2\sum_{k > j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \sin^2{\left(\frac{\dm_{kj}x}{4E}\right)}\,,
\end{align}
which is the probability of neutrino vacuum oscillations.
The calculation for antineutrinos yield the same result if one continues to assume the realness of the mixing matrix. 
In Fig.~\ref{fig:vac_osc}, we show the energy spectra of these oscillations in the low \si{\GeV} range for neutrinos that travel \SI{12000}{\km}\footnote{This baseline
is approximately the Earth diameter, and will be the maximum travel distance which we will use later on.}.
In the left-most panel, we see a supressed $\nm \to \ne$ transition. We can understand this behavior by studying Eq.~\ref{eq:Pab} and the 
numerical values of the mixing matrix in Eq.~\ref{eq:Uvalues}.
\si{\GeV} oscillations are driven by the $\dm[31]$ mass-squared difference. Hence, the most important mixing matrix 
elements for $P_{\a\b}$ are $U_{\a3}, U_{\a1}$ and $U_{\b3}, U_{\b1}$. The smallness of $U_{e3}$ then supresses the
amplitude of $\Pme$.
\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{figures/vac_osc.pdf}
    \caption{$\nm$ vacuum oscillations from Eq.~\ref{eq:Pab}}\label{fig:vac_osc}
\end{figure}

@@

We start with the effective Hamiltonian for the CC process. The Feynman rules for the left panel give us 
\begin{align}
    H_{CC} = \frac{G_F}{\sqrt{2}}\left[ \bar\ne \gamma^\rho\, (1- \gamma^5)\, e \right] \left[\bar e \,\gamma_\rho\, (1- \gamma^5)\, \ne \right]
\end{align}
By using the Fierz transformation 
\begin{align}
    \mathcal{L}^{V-A}(\psi_1,\psi_2,\psi_3,\psi_4) = \mathcal{L}^{V-A}(\psi_1,\psi_4,\psi_3,\psi_2)\,,
\end{align}
we can permute the terms inside the brackets, yielding
\begin{align}\label{eq:H_fierz}
    H_{CC} = \frac{G_F}{\sqrt{2}}\left[ \bar\ne \gamma^\rho\, (1- \gamma^5)\, \ne \right] \left[\bar e \,\gamma_\rho\, (1- \gamma^5)\, e \right]\,.
\end{align}
Now, lets consider a finite volume $V$ with electron states defined as 
\begin{align}\label{eq:e_states}
    \ket{e(p_e,h_e)} = \frac{1}{2E_eV}\, a_e^{(h_e)\dagger}(p_e)\ket{0}\,,
\end{align}
i.e. using the creation operator $a_e^{(h_e)\dagger}(p_e)$ to create electron states from vacuum with momenta $p_e$, energy $E_e$, and helicity $h_e$.
The density distribution of electrons in $V$ is $f(E_e,T)$, which we normalize to the total number of electrons as we integrate out the momenta $p_e$:
\begin{align}\label{eq:e_density}
    \int \dd p_e^3 f(E_e,T) = N_e V = n_e
\end{align}
Here, the electron density $N_e$ will ultimately determine the strength of the effective matter potential. 
To obtain the average effective Hamiltonian, project it on the electron states in Eq.~\ref{eq:e_states} and integrate over the density and sum over the helicities:

\begin{align}\label{eq:avg_H1}
    \avg{H_{CC}} &= \int \dd p_e^3\, \bra{e(p_e,h_e)} \times \frac{1}{2} \sum_{h_e} H   f(E_e,T) \ket{e(p_e,h_e)} \nonumber \\
           &= \frac{G_F}{\sqrt{2}} \int \dd p_e^3 \bra{e(p_e,h_e)} \left[ \bar\ne \gamma^\rho\, (1- \gamma^5)\, \ne \right]  f(E_e,T) \times \frac{1}{2} \sum_{h_e} \left[\bar e(x) \,\gamma_\rho\, (1- \gamma^5)\, e(x) \right] \ket{e(p_e,h_e)} \nonumber \\
           &= \frac{G_F}{\sqrt{2}} \bar\ne \gamma^\rho\, (1- \gamma^5)\, \ne \int \dd p_e^3  f(E_e,T) \times \frac{1}{2} \sum_{h_e} \bra{e(p_e,h_e)} \bar e(x) \,\gamma_\rho\, (1- \gamma^5)\, e(x)   \ket{e(p_e,h_e)}\,.
\end{align}
First, calculate the sum using trace technology
\begin{align}\label{eq:helicity_sum}
    \frac{1}{2} \sum_{h_e} \bra{e(p_e,h_e)} \bar e(x) \,\gamma_\rho\, (1- \gamma^5)\, e(x)   \ket{e(p_e,h_e)} &= \frac{1}{4E_e V} \sum_{h_e} \bar{u}_e^{h_e}(p_e) \,\gamma_\rho\, (1- \gamma^5)\, u_e^{h_e}(p_e) \nonumber \\
    &= \frac{1}{4 E_e V} \Trace{\left[ \sum_{h_e} \bar{u}_e^{h_e}(p_e) u_e^{h_e}(p_e) \,\gamma_\rho\, (1- \gamma^5)\, \right]} \nonumber \\
    &= \frac{1}{4 E_e V} \Trace{\left[ (\slashed{p}_e + m_e ) \,\gamma_\rho\, (1- \gamma^5)\, \right]} \nonumber \\
    &= \frac{(p_e)_\rho}{E_e V}\,.
\end{align}
Eq.~\ref{eq:avg_H1} now becomes 
\begin{align}\label{eq:avg_H2}
    \avg{H_{CC}} &= \frac{G_F}{\sqrt{2}E_e V} \bar\ne \, (1- \gamma^5)\, \ne \int \dd p_e^3\, \slashed{p}_e\, f(E_e,T)\,.
\end{align}
Expand the integral, and use the fact that $\vec{p}_e$ is odd:
\begin{align}
    \int \dd p_e^3\, \slashed{p}_e\, f(E_e,T) &= \int \dd p_e^3\, f(E_e,T) (\gamma^0 E_e - \vec{p}_e \cdot \vec{\gamma}) \nonumber \\
                                              &= \int \dd p_e^3\, f(E_e,T) \gamma^0 E_e \nonumber \\
                                              &= \gamma_0 E_e N_e V\,.
\end{align}
Inserting this into Eq.~\ref{eq:avg_H2}, we have
\begin{align}
    \avg{H_{CC}} &= \frac{G_F N_e}{\sqrt{2}} \bar{\nu}_e \, (1- \gamma^5)\, \ne \gamma_0 \nonumber \\
            &= \sqrt{2}G_F N_e \bar{\nu}_{eE} \gamma^0 \nu_{eL}\,,
\end{align}
where the projection operator $(1- \gamma^5)$ in the first line ensures that only the left-hand compontent of the neutrino fields interact. Thus,
\begin{align}\label{eq:V_CC1}
    V_{CC} = \sqrt{2}G_F N_e\,, \quad H_{CC}\ket{\nu_k} = V_{CC}\ket{\nu_k}\,.
\end{align}
Here we see a crucial difference between the eigenvectors between the vacuum Hamiltonian defined in Eq.~\ref{eq:osc_1} and $H_{CC}$, namely
that the CC interactions happen in the flavor basis rather than in the mass basis\footnote{A similar calculation for the NC reveals the same fact:
neutrinos interact in the flavor basis.}. In other words,
neutrinos propagate in their mass eigenstates, but interact in their flavor eigenstate. The mixing of mass eigenstates during propagation
determines if the flavor eigenstate has oscillated or not. The interactions still conserve lepton number, but the kinetic Lagrangian will not.

@@

\subsection{Matter Oscillations}
Since only $\ne$ undergo CC interactions in Earth-like matter, the $V_{CC}$ potential is zero for all other flavors. However, since all flavors undergo NC interactions the total matter potential in matrix form is
\begin{align}\label{eq:V_matrix}
    V = \begin{bmatrix}
        V_{CC} + V_{NC} & 0 & 0 \\
        0 & V_{NC} & 0 \\
        0 & 0 & V_{NC} \\
    \end{bmatrix} = V_{CC}\, \delta_{\alpha e}\mathbb{1} + V_{NC}\mathbb{1}\,.
\end{align}
Just as in Eq.~\ref{eq:TDSE}, we start with a Hamiltonian that solves the time-dependent Schrödinger equation. This time, let the Hamiltonian be 
\begin{align}
    H = H_0 + H_{I}\,,
\end{align}
where $H_0$ is the Hamiltonian in vacuum, and $H_{I}$ is our interaction Hamiltonian associated with our matter potentials.
Let the wavefunction that describes the $\nu_\a \to \nu_\b$ transition be denoted as
\begin{align}
    \psi_{\a\b}(t) = \bra{\nu_\b}\ket{\nu_\a (t)}\,,
\end{align}
i.e.~the evolution of the state of a neutrino emitted at $t =0$ with flavor $\alpha$ to flavor $\beta$ at time $t$.
